---
title: "Example of How to Retrieve Metadata from SRAdb "
author: "Candace Savonen"
date: "2020"
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Summary:

Sequence Read Archive (SRA) is a public repository of sequencing that has an R package, [SRAdb](https://rdrr.io/bioc/SRAdb/), on Bioconductor that can help you retrieve metadata associated with RNA-seq data. 
In this example, we will show you how to retrieve metadata from SRAdb for a project id (eg. SRP123456).
SRAdb has an [sqlite file](https://www.sqlite.org/aff_short.html) with all the information in it which we will download in this notebook. 
The tidyverse has the [`dbplyr` package](https://dbplyr.tidyverse.org/) made for handling `sqlite` files from R. 
You will recognize a lot of the functions are similar to what we use with data frames. 

### Set Up

For this example, we will use this SRA project id, [SRP045496](https://www.refine.bio/experiments/SRP045496/the-g-protein-alpha-subunit-gsa-is-a-tumor-suppressor-in-sonic-hedgehog-driven-medulloblastoma-rna-seq), which is for a medulloblastoma mouse

For more on SRA ids and more than you'll need to know about the SRA database, see this [knowledge base](https://www.ncbi.nlm.nih.gov/books/NBK56913/).

```{r}
# Declare the project id you are interested in.
study_id <- "SRP045496"
```

Install `SRAdb` if it is not installed.

```{r}
if (!("SRAdb" %in% installed.packages())) {
  BiocManager::install("SRAdb")
}
```

```{r}
# magrittr pipe
`%>%` <- dplyr::`%>%`
```

Set up SRA input directory and file path. 

```{r}
# Declare SRA directory 
sra_db_dir <- file.path("..", "..", "shared", "data", "SRAdb")

# Declare database file path
sqlite_file <- file.path(sra_db_dir, "SRAmetadb.sqlite")

if (!dir.exists(sra_db_dir)) {
  dir.create(sra_db_dir, recursive = TRUE)
}
```

Set up output directory and file path. 

```{r}
# Declare output directory 
output_dir <- file.path("training-modules", "SRA_metadata")

# Declare output file path using the project ID
output_metadata_file <- file.path(output_dir, paste0(study_id, "_metadata.tsv"))

# Create the directory if it doesn't exist.
if (!dir.exists(output_dir)) {
  dir.create(output_dir,  recursive = TRUE)
}
```

## Download and connect to SRA metadata sqlite file

Downloading this file will take some time, so we will only download it if it doesn't exist.

```{r}
# Check if the sqlite file exists here
if (!file.exists(sqlite_file)) {
  # Use SRAdb's function to download the file
  SRAdb::getSRAdbFile(destdir = sra_db_dir)
}
```

Connect to the sqlite file. 

```{r}
# Make connection to sqlite database file
sra_con <- DBI::dbConnect(RSQLite::SQLite(), sqlite_file)
```

## Find information on our selected SRA project id 

Use `dbplyr` functions to collect information related to our declared `study_id`.

```{r}
# Use the sqlite connection to connect to this `sra` table
sra_df <- dplyr::tbl(sra_con, "sra") %>% 
  # Filter to the study_id we are looking for
  dplyr::filter(study_accession == study_id) %>% 
  # Inner join to the study table that has more specific info about the study
  dplyr::inner_join(dplyr::tbl(sra_con, "study") , by = "study_accession") %>% 
  # We need to do this so the dbplyr queries are transformed to a dataset
  as.data.frame() 
```

## Retrieve sample-level information

Create a vector of sample ids that are related to our `study_id`.

```{r}
# Pull out sample accessions for the corresponding study
sample_accessions <- sra_df %>% 
  dplyr::pull(sample_accession)
```

Use `dbplyr` functions to filter to only the samples related to our `study_id`.

```{r}
# Connect to a different table in the sqlite file
sample_df <- dplyr::tbl(sra_con, "sample") %>% 
  # Collect the samples that we identified in the previous set of steps
  dplyr::filter(sample_accession %in% sample_accessions) %>%
  # Turn into a data.frame
  as.data.frame()
```

## Do some cleaning of the sample_df

Here we'll do some cleaning these cleaning steps will be dependent the metadata itself and on what information you are interested in.

```{r}
cleaned_sample_df <- sample_df %>% 
  dplyr::select(-which(apply(is.na(.), 2, all)))
```

## Write the sample_df to TSV file

```{r}
cleaned_sample_df %>% 
  readr::write_tsv(file.path(output_metadata_file))
```

### Print out session info

```{r}
sessionInfo()
```
